HLoRA: Hierarchical Bayesian Low-Rank Adaption for Multi-task Large Language Models

### 2024-06-06 paper/main.tex
resize figures, tables and equations to fit

\subsection{The Hierarchical Bayesian LoRA model} \label{sec:BoRA}
In the hierarchical LoRA model, we introduce a LoRA parameter set for each task $d$, denoted as $\theta_d$, containing all the low-rank decomposition matricies $A_d$ and $B_d$ from each decomposed layer in the Large Language Model. The likelihood of the data $\mathcal{D}$ given the task parameters $\theta_{1:D}$ is then given by (combining equation \ref{eq:data} and \ref{eq:LLMprob}):
\begin{equation} \label{eq:lik}
    \text{L}(\mathcal{D} | \theta_1, \theta_2, \ldots, \theta_D) = \prod_{d=1}^D L(\mathcal{D}_d | \theta_d) = \prod_{d=1}^D \prod_{n=1}^{N_d} \prod_{i=1}^{W_n-1} \text{LLM}(w_{d,n,i+1} | w_{d,n,1:i}; \theta_d).
\end{equation}
%
%
Tasks will share some knowledge with each other, and we model this by introducing a Gaussian hierarchical prior over the task parameters $\theta_{1:D}$, given by
\begin{equation} \label{eq:prior}
    P(\theta_{1:D} | \Theta, \tau ) = \prod_{d=1}^D N(\theta_d ; \Theta, \frac{1}{\tau} I)
\end{equation}
%
where $\Theta$ is a set of hierarchical mean parameters, $I$ is a unit diagonal matrix and $\tau \geq 0$ is a scalar precision hyperparameter controlling how similar the task parameters $\theta_d$ should be to the hierarchical mean parameters $\Theta$.
We let $\Theta$ have a uniform hyperprior (i.e. $P(\Theta) = 1$). The prior assumes that the task parameters share the common origin from the shared Gaussian prior, because the tasks are related: input data come from similar distributions and the output tasks has similar aspects. 
The precision hyperparameter $\tau$ controls how much the task parameters $\theta_d$ are a priori expected to be allowed to deviate from the hierarchical mean parameters $\Theta$. 
This is essentially a measure for how much structure and information the tasks share with each other, and should be set by the practitioner based on her's prior knowledge.

@@

\section{Experimental setup}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Use the Talk of Norway Dataset consisting of speeches of Norwegian parliament politicians. We will consider different speakers of this dataset as different tasks, and see how the hierarchical model can benefit from sharing information between these tasks.
% add figure of the dataset distribution in file training_lengths.png
\begin{figure}[h] \label{fig:task_len}
    \centering
    \includegraphics[width=\textwidth]{figures/training_lengths.png}
    \caption{Number of training examples for each task}
   \label{fig:training_lengths}
\end{figure}
% As base model, use opt-350m as it is predominantly trained on English, and only has seen a small amount of Norwegian data from the commoncrawls dataset.
To test our method, we optimize the model with a range of different precision hyperparameters $\tau$, and evaluate the models' Perplexity on the test set for each task.
% DEF PERPLEXITY
Perplexity is defined as the exponentiated average negative log-likelihood per token, and is a common metric for evaluating language models. A lower perplexity indicates a better model.
We use the Talk of Norway Dataset~\cite{lapponi_talk_2018}, which consists of speeches of Norwegian parliament politicians.
Each speaker will come from different political parties and geographical areas, but the tasks will still share a common domain. 
Therefore, we can consider different speakers of this dataset as different tasks. Specifically, we select chronologically the 25 first speakers that have above 100 speeches in the dataset.
This gives us a total of 25 tasks with samples ranging from 110 to 477 documents, and an average of 202 documents per task. For some speakers we have significantly less data as for others. 
The distibution of samples per task can be seen in Figure~\ref{fig:task_len}.
We leave 33\% of the data for the test set.

@@

\begin{table}[h] 
    \centering
    \begin{tabular}{ccc}
        \toprule
        Learning Rate & Precision ($\tau$) & Test Perplexity \\
        \midrule
            $10^{-4}$ &              0 &                 16.80 \\
            $10^{-5}$ &       $10^{0}$ &                 16.59 \\
            $10^{-4}$ &       $10^{1}$ &                 12.85 \\
            $10^{-3}$ &       $10^{2}$ &                 \textbf{12.82} \\
            $10^{-2}$ &       $10^{3}$ &                 13.26 \\
            $10^{-1}$ &       $10^{4}$ &                 13.91 \\
        \bottomrule
        \end{tabular}
        \caption{Empirical results of the hierarchical LORA finetuned on the three tasks. The perplexity is calculated on the test set.}
        \label{tbl:results}
    \end{table}

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/results_plot.png}
    \caption{Plot of Precision vs Perplexity. The thick black line is the overall test perplexity over all tasks, and the thinner grey lines are the test perplexity for each individual task.}
    \label{fig:results_plot}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/relative_improvement_combined.png}
    \caption{Task dataset size versus the relative improvement in perplexity of the best performing hierarchical model ($\tau = 100$) compared to the case when all models are trained independently ($\tau = 0$) in blue, and compared to the limiting case when all task specific model parameters are constraint to be equal ($\tau=10000$) in red. 
    The figure shows that all tasks benefit from sharing parameters with the global model, and indicates that the tasks with less data benefit more than the tasks with more data.}
    \label{fig:relative_improvement}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/trainingsize_vs_perplexity.png}
    \caption{Training size vs perplexity of the best performing hierarchical model ($\tau = 100$). The figure shows that there is no strong relationship between dataset size and final perplexity.}
    \label{fig:absolute_improvement}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/weights_vs_data_lengths.png}
    \caption{The figure shows the L2-distance between each task's adapter weight and the global prior on the y-axis, and the number of training examples on the x-axis. We see that, as expected, tasks with more training data will have a larger distance to the global prior.}
    \label{fig:weights_vs_datalen} 
\end{figure}

@@

\begin{table}[ht] \label{tbl:task_perplexities}
    \centering
    \caption{Test perplexity for each task and precision hyperparameter $\tau$}
    \begin{tabular}{lrrrrrr}
        \toprule
        tau & 0     & 1     & 10    &          100   & 1000  & 10000 \\
        Task                   &       &       &       &                &       &       \\
        \midrule
        abid q raja            & 15.56 & 15.20 & 12.57 & \textbf{11.60} & 11.91 & 12.41 \\
        akhtar chaudhry        & 14.78 & 14.37 & 12.01 & \textbf{11.43} & 11.66 & 12.06 \\
        aksel hagen            & 15.88 & 15.30 & 12.68 & \textbf{11.84} & 12.26 & 12.61 \\
        alf egil holmelid      & 16.96 & 16.86 & 13.97 & \textbf{12.66} & 13.35 & 14.51 \\
        anders b werp          & 16.86 & 16.35 & 13.04 & \textbf{11.87} & 12.54 & 13.00 \\
        andré n skjelstad      & 15.53 & 15.23 & 12.59 & \textbf{11.58} & 11.98 & 12.41 \\
        andré oktay dahl       & 16.01 & 15.85 & 13.50 & \textbf{12.55} & 12.88 & 13.48 \\
        anette trettebergstuen & 15.98 & 15.73 & 13.43 & \textbf{12.57} & 12.90 & 13.28 \\
        anita apelthun sæle    & 25.44 & 25.42 & 21.04 & \textbf{19.21} & 20.16 & 21.37 \\
        anne tingelstad wøien  & 16.41 & 16.23 & 13.60 & \textbf{12.43} & 12.90 & 13.54 \\
        anniken huitfeldt      & 14.44 & 14.39 & 12.33 & \textbf{11.66} & 12.05 & 12.53 \\
        ansgar gabrielsen      & 16.10 & 16.16 & 14.07 & \textbf{13.23} & 13.68 & 14.19 \\
        arild grande           & 16.65 & 16.29 & 13.31 & \textbf{12.21} & 12.56 & 13.08 \\
        arne lyngstad          & 15.57 & 15.28 & 12.12 & \textbf{11.25} & 11.66 & 12.08 \\
        arve kambe             & 17.78 & 17.40 & 14.24 & \textbf{13.11} & 13.43 & 13.79 \\
        asmund kristoffersen   & 17.15 & 16.75 & 13.63 & \textbf{12.61} & 13.03 & 13.61 \\
        audun lysbakken        & 15.21 & 15.25 & 13.22 & \textbf{12.48} & 12.67 & 13.13 \\
        bendiks h arnesen      & 14.34 & 14.23 & 12.04 & \textbf{11.11} & 11.42 & 11.97 \\
        bente thorsen          & 16.35 & 15.88 & 12.87 & \textbf{11.60} & 12.28 & 13.06 \\
        bjørg tørresdal        & 16.02 & 15.65 & 12.78 & \textbf{11.58} & 11.92 & 12.46 \\
        bjørn hernæs           & 16.33 & 15.94 & 13.68 & \textbf{12.76} & 13.08 & 13.58 \\
        bjørn jacobsen         & 22.28 & 22.87 & 19.54 & \textbf{18.52} & 19.05 & 20.53 \\
        bjørn lødemel          & 22.91 & 24.09 & 19.65 & \textbf{18.08} & 19.39 & 21.08 \\
        bjørn tore godal       & 16.57 & 16.30 & 14.55 & \textbf{14.02} & 14.22 & 14.48 \\
        borghild tenden        & 14.42 & 14.84 & 12.12 & \textbf{11.10} & 11.68 & 12.41 \\
        \bottomrule
    \end{tabular}
\end{table}

>>>>

\subsection{The Hierarchical Bayesian LoRA model} \label{sec:BoRA}
In the hierarchical LoRA model, we introduce a LoRA parameter set for each task $d$, denoted as $\theta_d$, containing all the low-rank decomposition matricies $A_d$ and $B_d$ from each decomposed layer in the Large Language Model. The likelihood of the data $\mathcal{D}$ given the task parameters $\theta_{1:D}$ is then given by (combining equation \ref{eq:data} and \ref{eq:LLMprob}):

\begin{align}  \label{eq:lik}
    & \text{L}(\mathcal{D} | \theta_1, \theta_2, \ldots, \theta_D) = \prod_{d=1}^D L(\mathcal{D}_d | \theta_d) \nonumber \\
    &= \prod_{d=1}^D \prod_{n=1}^{N_d} \prod_{i=1}^{W_n-1} \text{LLM}(w_{d,n,i+1} | w_{d,n,1:i}; \theta_d).
\end{align}
%
%
Tasks will share some knowledge with each other, and we model this by introducing a Gaussian hierarchical prior over the task parameters $\theta_{1:D}$, given by
\begin{equation} \label{eq:prior}
    P(\theta_{1:D} | \Theta, \tau ) = \prod_{d=1}^D N(\theta_d ; \Theta, \frac{1}{\tau} I)
\end{equation}
%
where $\Theta$ is a set of hierarchical mean parameters, $I$ is a unit diagonal matrix and $\tau \geq 0$ is a scalar precision hyperparameter controlling how similar the task parameters $\theta_d$ should be to the hierarchical mean parameters $\Theta$.
We let $\Theta$ have a uniform hyperprior (i.e. $P(\Theta) = 1$). The prior assumes that the task parameters share the common origin from the shared Gaussian prior, because the tasks are related: input data come from similar distributions and the output tasks has similar aspects. 
The precision hyperparameter $\tau$ controls how much the task parameters $\theta_d$ are a priori expected to be allowed to deviate from the hierarchical mean parameters $\Theta$. 
This is essentially a measure for how much structure and information the tasks share with each other, and should be set by the practitioner based on her's prior knowledge.

@@

\section{Experimental setup}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Use the Talk of Norway Dataset consisting of speeches of Norwegian parliament politicians. We will consider different speakers of this dataset as different tasks, and see how the hierarchical model can benefit from sharing information between these tasks.
% add figure of the dataset distribution in file training_lengths.png
\begin{figure} \label{fig:task_len}
    \centering
    \resizebox{\linewidth}{!}{%
    \includegraphics[width=\textwidth]{figures/training_lengths.png}
    }
    \caption{Number of training examples for each task}
\label{fig:training_lengths}
\end{figure}
% As base model, use opt-350m as it is predominantly trained on English, and only has seen a small amount of Norwegian data from the commoncrawls dataset.
To test our method, we optimize the model with a range of different precision hyperparameters $\tau$, and evaluate the models' Perplexity on the test set for each task.
% DEF PERPLEXITY
Perplexity is defined as the exponentiated average negative log-likelihood per token, and is a common metric for evaluating language models. A lower perplexity indicates a better model.
We use the Talk of Norway Dataset~\cite{lapponi_talk_2018}, which consists of speeches of Norwegian parliament politicians.
Each speaker will come from different political parties and geographical areas, but the tasks will still share a common domain. 
Therefore, we can consider different speakers of this dataset as different tasks. Specifically, we select chronologically the 25 first speakers that have above 100 speeches in the dataset.
This gives us a total of 25 tasks with samples ranging from 110 to 477 documents, and an average of 202 documents per task. For some speakers we have significantly less data as for others. 
The distibution of samples per task can be seen in Figure~\ref{fig:task_len}.
We leave 33\% of the data for the test set.

@@

\begin{lstlisting}[gobble=2]
    \begin{table}[h] 
        \centering
        \resizebox{\linewidth}{!}{%
            \begin{tabular}{ccc}
                \toprule
                Learning Rate & Precision ($\tau$) & Test Perplexity \\
                \midrule
                    $10^{-4}$ &              0 &                 16.80 \\
                    $10^{-5}$ &       $10^{0}$ &                 16.59 \\
                    $10^{-4}$ &       $10^{1}$ &                 12.85 \\
                    $10^{-3}$ &       $10^{2}$ &                 \textbf{12.82} \\
                    $10^{-2}$ &       $10^{3}$ &                 13.26 \\
                    $10^{-1}$ &       $10^{4}$ &                 13.91 \\
                \bottomrule
                \end{tabular}
        }
            \caption{Empirical results of the hierarchical LORA finetuned on the three tasks. The perplexity is calculated on the test set.}
            \label{tbl:results}
        \end{table}
\end{lstlisting}

\begin{figure}%[ht]
    %\centering
    \resizebox{\linewidth}{!}{%
    \includegraphics{figures/results_plot.png}
    }
    \caption{Plot of Precision vs Perplexity. The thick black line is the overall test perplexity over all tasks, and the thinner grey lines are the test perplexity for each individual task.}
    \label{fig:results_plot}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \resizebox{\linewidth}{!}{%
    \includegraphics[width=\textwidth]{figures/relative_improvement_combined.png}
    }
    \caption{Task dataset size versus the relative improvement in perplexity of the best performing hierarchical model ($\tau = 100$) compared to the case when all models are trained independently ($\tau = 0$) in blue, and compared to the limiting case when all task specific model parameters are constraint to be equal ($\tau=10000$) in red. 
    The figure shows that all tasks benefit from sharing parameters with the global model, and indicates that the tasks with less data benefit more than the tasks with more data.}
    \label{fig:relative_improvement}
\end{figure}

@@

\begin{figure}
    \centering
    \resizebox{\linewidth}{!}{%
    \includegraphics[width=\textwidth]{figures/trainingsize_vs_perplexity.png}
    }
    \caption{Training size vs perplexity of the best performing hierarchical model ($\tau = 100$). The figure shows that there is no strong relationship between dataset size and final perplexity.}
    \label{fig:absolute_improvement}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \resizebox{\linewidth}{!}{%
    \includegraphics[width=\textwidth]{figures/weights_vs_data_lengths.png}
    }
    \caption{The figure shows the L2-distance between each task's adapter weight and the global prior on the y-axis, and the number of training examples on the x-axis. We see that, as expected, tasks with more training data will have a larger distance to the global prior.}
    \label{fig:weights_vs_datalen} 
\end{figure}

@@

\begin{table}[ht] \label{tbl:task_perplexities}
    \centering
    \caption{Test perplexity for each task and precision hyperparameter $\tau$}
    \resizebox{\linewidth}{!}{%
    \begin{tabular}{lrrrrrr}
        \toprule
        tau & 0     & 1     & 10    &          100   & 1000  & 10000 \\
        Task                   &       &       &       &                &       &       \\
        \midrule
        abid q raja            & 15.56 & 15.20 & 12.57 & \textbf{11.60} & 11.91 & 12.41 \\
        akhtar chaudhry        & 14.78 & 14.37 & 12.01 & \textbf{11.43} & 11.66 & 12.06 \\
        aksel hagen            & 15.88 & 15.30 & 12.68 & \textbf{11.84} & 12.26 & 12.61 \\
        alf egil holmelid      & 16.96 & 16.86 & 13.97 & \textbf{12.66} & 13.35 & 14.51 \\
        anders b werp          & 16.86 & 16.35 & 13.04 & \textbf{11.87} & 12.54 & 13.00 \\
        andré n skjelstad      & 15.53 & 15.23 & 12.59 & \textbf{11.58} & 11.98 & 12.41 \\
        andré oktay dahl       & 16.01 & 15.85 & 13.50 & \textbf{12.55} & 12.88 & 13.48 \\
        anette trettebergstuen & 15.98 & 15.73 & 13.43 & \textbf{12.57} & 12.90 & 13.28 \\
        anita apelthun sæle    & 25.44 & 25.42 & 21.04 & \textbf{19.21} & 20.16 & 21.37 \\
        anne tingelstad wøien  & 16.41 & 16.23 & 13.60 & \textbf{12.43} & 12.90 & 13.54 \\
        anniken huitfeldt      & 14.44 & 14.39 & 12.33 & \textbf{11.66} & 12.05 & 12.53 \\
        ansgar gabrielsen      & 16.10 & 16.16 & 14.07 & \textbf{13.23} & 13.68 & 14.19 \\
        arild grande           & 16.65 & 16.29 & 13.31 & \textbf{12.21} & 12.56 & 13.08 \\
        arne lyngstad          & 15.57 & 15.28 & 12.12 & \textbf{11.25} & 11.66 & 12.08 \\
        arve kambe             & 17.78 & 17.40 & 14.24 & \textbf{13.11} & 13.43 & 13.79 \\
        asmund kristoffersen   & 17.15 & 16.75 & 13.63 & \textbf{12.61} & 13.03 & 13.61 \\
        audun lysbakken        & 15.21 & 15.25 & 13.22 & \textbf{12.48} & 12.67 & 13.13 \\
        bendiks h arnesen      & 14.34 & 14.23 & 12.04 & \textbf{11.11} & 11.42 & 11.97 \\
        bente thorsen          & 16.35 & 15.88 & 12.87 & \textbf{11.60} & 12.28 & 13.06 \\
        bjørg tørresdal        & 16.02 & 15.65 & 12.78 & \textbf{11.58} & 11.92 & 12.46 \\
        bjørn hernæs           & 16.33 & 15.94 & 13.68 & \textbf{12.76} & 13.08 & 13.58 \\
        bjørn jacobsen         & 22.28 & 22.87 & 19.54 & \textbf{18.52} & 19.05 & 20.53 \\
        bjørn lødemel          & 22.91 & 24.09 & 19.65 & \textbf{18.08} & 19.39 & 21.08 \\
        bjørn tore godal       & 16.57 & 16.30 & 14.55 & \textbf{14.02} & 14.22 & 14.48 \\
        borghild tenden        & 14.42 & 14.84 & 12.12 & \textbf{11.10} & 11.68 & 12.41 \\
        \bottomrule
    \end{tabular}
    }
\end{table}
